---
title: "PANS results in Forestry Statistics Chapter 6"
author: "Daniel Braby, Forest Information and Statistics"
date: "2023-08-30"
output: html_document
editor_options: 
  chunk_output_type: console
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```

# Setup

```{r}
library(here)
library(haven)
library(labelled)
library(dplyr)
library(tidyr)
library(stringr)
library(lubridate)
```

```{r}
round2 = function(x, digits) {
  posneg = sign(x)
  z = abs(x)*10^digits
  z = z + 0.5 + sqrt(.Machine$double.eps)
  z = trunc(z)
  z = z/10^digits
  z*posneg
}
```

```{r}
pans_raw <- read_sav(here("raw-data",
                          "262400153 PANS SPSS Q1-11 Safeguarded 090523.sav"))
```

Convert all cells to their labels (except the columns which contains the word "Weight").

```{r}
pans_lbld <- pans_raw |> 
  mutate(across(.cols = !contains("Weight"), ~to_factor(.)))
```

# EDA

I think something unusual is going on with the `No_Of_Visits` variable. It's labelled and converting to numeric appears to be doing weird things. Explore this before continuing.

```{r}
pans_raw |> 
  select(No_Of_Visits) |> 
  mutate(label = to_factor(No_Of_Visits),
         label_numeric = as.numeric(No_Of_Visits)) |> 
  filter(!is.na(No_Of_Visits)) |> 
  filter(label != label_numeric)
```

From Oct-22 to Dec-22, there were [932,443 visits to green and natural spaces](https://www.gov.uk/government/statistics/the-people-and-nature-survey-for-england-adult-data-y3q3-october-2022-december-2022-official-statistics) in England. Try to reproduce this number.

```{r}
pans_raw |> 
  select(Wave, M2, No_Of_Visits, Weight_Grossed_No_Of_Visits) |> 
  filter(M2 == 1) |> 
  select(-M2) |> 
  mutate(Wave = to_factor(Wave),
         No_Of_Visits = remove_val_labels(No_Of_Visits)) |> 
  filter(No_Of_Visits >=0) |> 
  separate(col = Wave,
           into = c("wave", "date"),
           sep = " - ") |> 
  mutate(date = my(date)) |> 
  filter(date >= "2022-10-01" & date <= "2022-12-01") |> 
  select(-c(wave)) |> 
  mutate(across(-date, ~as.numeric(.x))) |> 
  mutate(nb_visit = abs(No_Of_Visits) * Weight_Grossed_No_Of_Visits) |> 
  summarise(nb_visit = sum(nb_visit, na.rm = TRUE))
```

From Ann:

> The total number of visits from April 2020 to end of Dec 2022 should be 11399746...let me know if you get a different number and I can double check. For the first two years, i.e., April 2020 to end of March 2022 I get 8521838.

```{r}
pans_raw |> 
  select(Wave, M2, No_Of_Visits, Weight_Grossed_No_Of_Visits) |> 
  filter(M2 == 1) |> 
  select(-M2) |> 
  mutate(Wave = to_factor(Wave),
         No_Of_Visits = remove_val_labels(No_Of_Visits)) |> 
  filter(No_Of_Visits >=0) |> 
  separate(col = Wave,
           into = c("wave", "date"),
           sep = " - ") |> 
  mutate(date = my(date)) |> 
  filter(date >= "2020-04-01" & date <= "2022-03-01") |> 
  select(-c(wave)) |> 
  mutate(across(-date, ~as.numeric(.x))) |> 
  mutate(nb_visit = abs(No_Of_Visits) * Weight_Grossed_No_Of_Visits) |> 
  summarise(nb_visit = sum(nb_visit, na.rm = TRUE))
```

Good news: can confirm both numbers!

# Proportion of respondents that visited woodland in the last month

Around 30% said visited forest or woodland in the last month.

```{r}
woodland_visit_pct <- pans_lbld |> 
  select(Wave, starts_with("M1_Q2"), Weight_Percent) |> 
  separate(col = Wave,
           into = c("wave", "date"),
           sep = " - ") |> 
  mutate(date = my(date),
         fy = nprs::date2fy(date, fye = TRUE)) |> 
  sjlabelled::label_to_colnames() |> 
  rename(weight = `Weight - percentages ONLY - Modules 1, 2, 3/4 & 6`) |> 
  rename_with(.cols = -c(wave, weight),
              ~str_remove_all(.x, "Which of the following type\\(s\\) of green and natural spaces have you visited during the last month: ")) |> 
  mutate(across(where(is.factor), ~as.character(.x))) |> 
  pivot_longer(cols = -c(wave, weight, date, fy),
               names_to = "space",
               values_to = "answer") |> 
  group_by(fy, space, answer) |> 
  summarise(weight = sum(weight, na.rm = TRUE)) |> 
  ungroup() |> 
  filter(!is.na(answer)) |> 
  group_by(fy, space) |> 
  mutate(pct = weight / sum(weight)) |> 
  ungroup() |> 
  filter(answer == "Yes",
         str_detect(space, "Woodland")) |> 
  select(fy, space, pct)
```

# Table 6.1

# Number of visits to forest and woodland (module 2): asked in last 2 weeks and grossed to English population in last month

The relevant columns to explore annual number of visits to forest and woodland are:

1. `Wave`: each wave corresponds to a month.
2. `M2A_Q2`: "Which of these best describes the main destination of your visit? (i.e. the place you spent most time in)" 
3. `No_Of_Visits`: "How many times, if at all, did you make this type of visit to green and natural spaces in the last 14 days?"
4. `Weight_Grossed_No_Of_Visits`: estimated total number of visits to green and natural spaces in the past month (in '000s).

## Total number of visits to green and natural spaces where forest or woodland was the main destination.

```{r}
pans_raw |> 
  select(Respondent_ID, Wave, M2A_Q2, No_Of_Visits, Weight_Grossed_No_Of_Visits) |> 
  filter(M2A_Q2 == 4) |> 
  select(-M2A_Q2) |> 
  mutate(Wave = to_factor(Wave)) |> 
  separate(col = Wave,
           into = c("wave", "date"),
           sep = " - ") |> 
  mutate(date = my(date),
         fy = nprs::date2fy(date, fye = TRUE)) |> 
  filter(No_Of_Visits > 0) |> # remove "Prefer not to say" and "Don't know" 
  select(-c(wave)) |> 
  mutate(No_Of_Visits_num = as.numeric(No_Of_Visits)) |> 
  mutate(nb_visit = No_Of_Visits_num * Weight_Grossed_No_Of_Visits) |> 
  group_by(fy) |> 
  summarise(nb_visit = sum(nb_visit, na.rm = TRUE)) |> 
  ungroup() |> 
  mutate(nb_visit = (nb_visit * 1e3) / 1e6)
```

# Table 6.2a

# Main mode of transport

This question was only asked to 30% of those that reported visiting woodland in the last 14 days.

1. `M2A_SUB_Q1`: What was the main form of transport you used to reach your main destination?

```{r}
transport_mode <- pans_lbld |> 
  select(Wave, M2A_Q2, M2A_SUB_Q1, Weight_Percent_M2A_SUB) |> 
  filter(M2A_Q2 == "Woodland or forest") |> 
  filter(!is.na(M2A_SUB_Q1)) |> 
  group_by(Wave, M2A_SUB_Q1) |> 
  summarise(weight = sum(Weight_Percent_M2A_SUB, na.rm = TRUE)) |> 
  ungroup() |> 
  separate(col = Wave,
           into = c("wave", "date"),
           sep = " - ") |> 
  mutate(date = my(date),
         fy = nprs::date2fy(date, fye = TRUE)) |> 
  filter(fy %in% c(2021, 2022)) |> 
  group_by(fy, M2A_SUB_Q1) |> 
  summarise(weight = sum(weight, na.rm = TRUE)) |> 
  ungroup() |> 
  group_by(fy) |> 
  mutate(pct = weight/sum(weight)) |> 
  ungroup()
```

```{r}
transport_mode |> 
  select(-weight) |> 
  pivot_wider(names_from = "fy",
              values_from = "pct") |> 
  arrange(desc(`2022`)) |> 
  mutate(across(where(is.numeric), ~round2(.x*100, 0)))
```

# Table 6.2b

M2A_Q6

```{r}
dist_travelled <- pans_lbld |> 
  select(Wave, M2A_Q2, M2A_Q6, Weight_Percent_M2A) |> 
  filter(M2A_Q2 == "Woodland or forest") |> 
  filter(!is.na(M2A_Q6)) |> 
  group_by(Wave, M2A_Q6) |> 
  summarise(weight = sum(Weight_Percent_M2A, na.rm = TRUE)) |> 
  ungroup() |> 
  separate(col = Wave,
           into = c("wave", "date"),
           sep = " - ") |> 
  mutate(date = my(date),
         fy = nprs::date2fy(date, fye = TRUE)) |> 
  filter(fy %in% c(2021, 2022)) |> 
  group_by(fy, M2A_Q6) |> 
  summarise(weight = sum(weight, na.rm = TRUE)) |> 
  ungroup() |> 
  group_by(fy) |> 
  mutate(pct = weight/sum(weight)) |> 
  ungroup()
```

```{r}
dist_travelled |> 
  select(-weight) |> 
  pivot_wider(names_from = "fy",
              values_from = "pct") |> 
  mutate(across(where(is.numeric), ~round2(.x*100, 0)))
```

# Table 6.2c

M2A_SUB_Q5: What were the main three reasons for taking this visit?

```{r}
motivation <- pans_lbld |> 
  select(Wave, M2A_Q2, contains("M2A_SUB_Q5"), Weight_Percent_M2A_SUB) |> 
  filter(M2A_Q2 == "Woodland or forest") |> 
  select(-M2A_Q2) |> 
  sjlabelled::label_to_colnames() |> 
  pivot_longer(cols = -c(`Month of interview`, `Weight - percentages ONLY - Module 2A_SUB (visit level analysis)`),
               names_to = "motivation",
               values_to = "yay_nay") |> 
  rename(Wave = `Month of interview`,
         Weight_Percent_M2A_SUB = `Weight - percentages ONLY - Module 2A_SUB (visit level analysis)`) |> 
  filter(yay_nay == "Yes") |> 
  select(-c(yay_nay)) |> 
  mutate(motivation = str_remove_all(motivation, "What was the main reason for taking this visit\\?\\: ")) |> 
  group_by(Wave, motivation) |> 
  summarise(weight = sum(Weight_Percent_M2A_SUB, na.rm = TRUE)) |> 
  ungroup() |> 
  separate(col = Wave,
           into = c("wave", "date"),
           sep = " - ") |> 
  mutate(date = my(date),
         fy = nprs::date2fy(date, fye = TRUE)) |> 
  filter(fy %in% c(2021, 2022)) |> 
  group_by(fy, motivation) |> 
  summarise(weight = sum(weight, na.rm = TRUE)) |> 
  ungroup() |> 
  group_by(fy) |> 
  mutate(pct = weight/sum(weight)) |> 
  ungroup()
```

```{r}
motivation <- pans_lbld |> 
  select(Wave, M2A_Q2, contains("M2A_SUB_Q5"), Weight_Percent_M2A_SUB) |> 
  filter(M2A_Q2 == "Woodland or forest") |> 
  select(-M2A_Q2) |> 
  sjlabelled::label_to_colnames() |> 
  pivot_longer(cols = -c(`Month of interview`,
                         `Weight - percentages ONLY - Module 2A_SUB (visit level analysis)`),
               names_to = "motivation",
               values_to = "yay_nay") |> 
  mutate(motivation = str_remove_all(motivation, "What was the main reason for taking this visit\\?\\: ")) |> 
  rename(Wave = `Month of interview`,
         Weight_Percent_M2A_SUB = `Weight - percentages ONLY - Module 2A_SUB (visit level analysis)`) |> 
  filter(!is.na(Weight_Percent_M2A_SUB )) |> 
  group_by(Wave, motivation, yay_nay) |> 
  summarise(weight = sum(Weight_Percent_M2A_SUB, na.rm = TRUE)) |> 
  ungroup() |> 
  separate(col = Wave,
           into = c("wave", "date"),
           sep = " - ") |> 
  mutate(date = my(date),
         fy = nprs::date2fy(date, fye = TRUE)) |> 
  filter(fy %in% c(2021, 2022)) |> 
  group_by(fy, motivation, yay_nay) |> 
  summarise(weight = sum(weight, na.rm = TRUE)) |> 
  ungroup() |> 
  group_by(fy, motivation) |> 
  mutate(pct = weight/sum(weight)) |> 
  ungroup() |> 
  filter(yay_nay == "Yes")
```

```{r}
motivation |> 
  select(-weight,
         -yay_nay) |> 
  pivot_wider(names_from = "fy",
              values_from = "pct") |> 
  arrange(desc(`2022`)) |> 
  mutate(across(where(is.numeric), ~round2(.x*100, 0)))
```

# Table 6.2d

M2A_Q8A

```{r}
visit_activity <- pans_lbld |> 
  select(Wave, M2A_Q2, contains("M2A_Q8A"), Weight_Percent_M2A) |> 
  filter(M2A_Q2 == "Woodland or forest") |> 
  select(-M2A_Q2) |> 
  sjlabelled::label_to_colnames() |> 
  pivot_longer(cols = -c(`Month of interview`, `Weight - percentages ONLY - Module 2A (visit level analysis)`),
               names_to = "activity",
               values_to = "yay_nay") |> 
  rename(Wave = `Month of interview`,
         Weight_Percent_M2A = `Weight - percentages ONLY - Module 2A (visit level analysis)`) |> 
  # filter(yay_nay == "Yes") |> 
  # select(-c(yay_nay)) |> 
  mutate(activity = str_remove_all(activity, "Which of these activities, if any, did you do on this specific visit\\? \\: ")) |> 
  group_by(Wave, activity, yay_nay) |> 
  summarise(weight = sum(Weight_Percent_M2A, na.rm = TRUE)) |> 
  ungroup() |> 
  separate(col = Wave,
           into = c("wave", "date"),
           sep = " - ") |> 
  mutate(date = my(date),
         fy = nprs::date2fy(date, fye = TRUE)) |> 
  filter(fy %in% c(2021, 2022)) |> 
  group_by(fy, activity, yay_nay) |> 
  summarise(weight = sum(weight, na.rm = TRUE)) |> 
  ungroup() |> 
  group_by(fy, activity) |> 
  mutate(pct = weight/sum(weight)) |> 
  ungroup() |> 
  filter(yay_nay == "Yes")
```

```{r}
visit_activity |> 
  select(-weight, -yay_nay) |> 
  pivot_wider(names_from = "fy",
              values_from = "pct") |> 
  arrange(desc(`2022`)) |> 
  mutate(across(where(is.numeric), ~round2(.x*100, 0)))
```

# Table 6.2e

M2A_Q7

```{r}
visit_duration <- pans_lbld |> 
  select(Wave, M2A_Q2, M2A_Q7, Weight_Percent_M2A) |> 
  filter(M2A_Q2 == "Woodland or forest") |> 
  filter(!is.na(M2A_Q7)) |> 
  group_by(Wave, M2A_Q7) |> 
  summarise(weight = sum(Weight_Percent_M2A, na.rm = TRUE)) |> 
  ungroup() |> 
  separate(col = Wave,
           into = c("wave", "date"),
           sep = " - ") |> 
  mutate(date = my(date),
         fy = nprs::date2fy(date, fye = TRUE)) |> 
  filter(fy %in% c(2021, 2022)) |> 
  group_by(fy, M2A_Q7) |> 
  summarise(weight = sum(weight, na.rm = TRUE)) |> 
  ungroup() |> 
  group_by(fy) |> 
  mutate(pct = weight/sum(weight)) |> 
  ungroup()
```

```{r}
visit_duration |> 
  select(-weight) |> 
  pivot_wider(names_from = "fy",
              values_from = "pct") |> 
  mutate(across(where(is.numeric), ~round2(.x*100, 0)))
```
