---
title: "An analysis of woodland engagement and socio-demographic profile in England's People and Nature Survey (2020/21)"
title-block-banner: "#6c207f"
author:
  - name: Si Maxwell
    affiliations:
      - Forest Information and Statistics
description: "Programme 3: Societal benefits of trees, woods and forests"
date: today
date-format: "D MMMM YYYY"
embed-resources: TRUE
format:
  docx:
    reference-doc: Z:\IFOS\Statistics\Data\Social\Programme 3 work\programme3\fr-reference-doc.docx
editor: visual
editor_options: 
  chunk_output_type: console
---

```{r}
#| include: false
# 
if(!require(pacman)) {install.packages("pacman")}
pacman::p_load(here,
               readr,
               readxl,
               dplyr,
               tidyr,
               forcats,
               ggplot2,
               scales,
               ggeffects,
               performance)
```

```{r}
#| include: false
full_model <- read_rds(here("data", "full-model.rds"))
```

```{r}
#| include: false
source(here("src", "colour_selector.R"))
source(here("src", "scale_gss.R"))
source(here("src", "theme_gss.R"))
```

# Methodology

## Study design and sample

Data were from year 1 of the [People and Nature Survey](https://www.gov.uk/government/collections/people-and-nature-survey-for-england) (PANS) for England (Natural England, 2022). PANS is a nationally representative omnibus survey commissioned by Natural England which gathers information on people's engagement, understanding and attitude towards the natural environment. Data were collected online (by Kantar). The survey builds on the Monitor of Engagement with the Natural Environment (MENE) survey which ran from 2009 to 2019.

## Data

Woodland engagement was estimated by asking *Which of the following type(s) of green and natural spaces have you visited during the last month?* (M1_Q2_4). Of the 24,994 respondents, 24,017 (96%) answered yes or no to M1_Q2_4, with 67% reporting they visited woodland or forest (yes), 29% reporting they not visit woodland or forest (no), and 4% non-response (opt-outs).

The following socio-demographic characteristics were considered in the analysis: age, education (level of highest qualification), employment status, ethnicity, gender, Index of Multiple Deprivation (IMD) decile, income, home region, and urban-rural classification. Socio-demographic characteristics are summarised in Table 1.

#### Table 1: Breakdown of socio-demographic profile

```{r}
#| echo: false
read_xlsx(here("pans-report", "year1", "pans-table1.xlsx")) %>% 
  mutate(`Socio-demographic characteristic` = replace_na(`Socio-demographic characteristic`, "")) %>% 
  knitr::kable()
```

## Statistical analysis

All analyses were conducted in R 4.2.1 (R Core Team, 2022). We used generalised linear modelling to investigate the association between woodland engagement (dependent/response variable) and socio-demographic profile (independent variables/predictors).

Models were created for all possible combination of predictors plus an intercept-only null model, representing statistically random variation around a constant mean. An information theoretic approach was taken, with model performance assessed using Akaike's Information Criterion (AIC), Akaike weights ($w_i$) and Tjur's Coefficient of Discrimination ($R^2$). Collinearity occurs when predictors are correlated with one another, and leads to over-estimation of model fit. The variance inflation factor (VIF) was used to measure collinearity, with a cut-off of VIF \< 5.

We excluded urban/rural classification from the analysis for two reasons. First, respondents were asked to self-define their urban/rural classification, therefore, answers were subjective and did not follow the [2011 Urban/Rural Classification scheme for official statistics](https://www.gov.uk/government/collections/rural-urban-classification). Second, for respondents who completed the urban/rural question, their IMD decile was not reported. With 4,774 respondents reporting an urban/rural classification and 20,082 respondents reporting an IMD decile (see @sec-supplementary-table-1), we chose to maximise sample size and exclude urban/rural classification. However, it should be noted that a model which included age, education, employment status, ethnicity, gender, income, home region, and urban/rural classification was overwhelmingly supported (AIC = 4,054.7, $w_i$ \> 0.999) compared with a model replacing urban/rural classification with IMD decile (AIC = 20,207.1, $w_i$ \< 0.01) (see @sec-supplementary-table-2).

Significant socio-demographic characteristics were identified based on Analysis of Variance (ANOVA) using likelihood ratio chi-squared statistics (Fox & Weisberg, 2019). Estimated marginal means for each socio-demographic characteristic were estimated post-hoc using the `ggemmeans()` function in the `{ggeffects}` package (Lüdecke, 2018). Model fit statistics (AIC, $w_i$, Tjur's $R^2$) were calculated using the `{performance}` package (Lüdecke, 2021).

# Results

A model in which woodland engagement was regressed against all predictors (age, employment status, ethnicity, gender, IMD decile, income, qualification level and home region) was the best-supported model ($w_i$ = 0.85), with socio-demographic characteristics explaining approximately 5% (Tjur's $R^2$ = 0.05) of the variance in woodland engagement. Summary statistics for all models with $w_i > 0.01$ can be found in Supplementary Table 3. The findings presented here are for the best-supported full model only.

All socio-demographic characteristics were statistically significant at $p < 0.01$ except for education.

Woodland engagement was higher among younger people than older people. For example, in the 16-24 age group, an estimated 33% (95% CI: 30-37%) of people reported engaging with woodland in the last month, compared with an estimated 19% (95% CI: 17-22%) of people aged 65 and over (Figure 1). These results are consistent with the addition of long-term illness as a binary predictor and an age-illnesses interaction term.

#### Figure 1: Estimated marginal means for woodland engagement by age

```{r}
#| echo: false
ggemmeans(full_model, "age", rg.limit = 72000) %>% 
  as.data.frame() %>% 
  ggplot(aes(x, predicted)) +
  geom_point(size = 2,
             colour = colour_selector(1)) +
  geom_linerange(aes(ymin = conf.low,
                     ymax = conf.high),
                 linewidth = 1,
                 colour = colour_selector(1)) +
  scale_y_cont_gss(labels = percent,
                   breaks = seq(0, 1, 0.2),
                   limits = c(0, 1)) +
  labs(x = NULL,
       y = NULL,
       subtitle = "Estimated proportion of woodland visitors") +
  theme_gss(tick_marks = T)
```

People from the South East and South West are more likely to report having visited woodland in the last month, compared to the other regions of England (Figure 2). People from London were least likely to have visited woodland in the last month (estimated marginal mean of 20% (95% CI: 17-22%).

#### Figure 2: Estimated marginal means for woodland engagement by home region

```{r}
#| echo: false
ggemmeans(full_model, "region", rg.limit = 36000) %>% 
  as.data.frame() %>% 
  as_tibble() %>% 
  mutate(x = fct_reorder(x, predicted)) %>% 
  ggplot(aes(predicted, x)) +
  geom_point(size = 2,
             colour = colour_selector(1)) +
  geom_linerange(aes(xmin = conf.low,
                     xmax = conf.high),
                 linewidth = 1,
                 colour = colour_selector(1)) +
  scale_x_cont_gss(labels = percent,
                   breaks = seq(0, 1, 0.2),
                   limits = c(0, 1)) +
  labs(x = NULL,
       y = NULL,
       subtitle = "Estimated proportion of woodland visitors") +
  theme_gss(grid = "x")
```

An estimated 21% (95% CI: 19-23%) of people from IMD decile 1 (most deprived) areas visited woodland in the last month, compared to an estimated 31% (95% CI: 28-34%) from IMD decile 10 (least deprived) areas. As with age, a consistent, monotonic trend is observed with woodland engagement and IMD decile, where woodland engagement increases with decreasing deprivation.

#### Figure 3: Estimated marginal means for woodland engagement by home Index of Multiple Deprivation (IMD) decile

```{r}
#| echo: false
ggemmeans(full_model, "imd_decile", rg.limit = 360000) %>% 
  as.data.frame() %>% 
  ggplot(aes(x, predicted)) +
  geom_point(size = 2,
             colour = colour_selector(1)) +
  geom_linerange(aes(ymin = conf.low,
                     ymax = conf.high),
                 linewidth = 1,
                 colour = colour_selector(1)) +
  scale_x_cont_gss(breaks = seq(1, 10, 1),
                   limits = c(1, 10)) +
  scale_y_cont_gss(labels = percent,
                   breaks = seq(0, 1, 0.2),
                   limits = c(0, 1)) +
  labs(x = NULL,
       y = NULL,
       subtitle = "Estimated proportion of woodland visitors") +
  theme_gss(grid = "y",
            tick_marks = T)
```

Higher income people are more likely to engage with woodland or forest compared with lower income people. For example, in the £50,000+ income band, an estimated 30% (95% CI: 27-33%) of people visited woodland in the last month, compared to 22% (95% CI: 19-24) in the £0-14,999.

#### Figure 4: Estimated marginal means for woodland engagement by income

```{r}
#| echo: false
ggemmeans(full_model, "income", rg.limit = 36000) %>% 
  as.data.frame() %>% 
  ggplot(aes(predicted, x)) +
  geom_point(size = 2,
             colour = colour_selector(1)) +
  geom_linerange(aes(xmin = conf.low,
                     xmax = conf.high),
                 linewidth = 1,
                 colour = colour_selector(1)) +
  scale_x_cont_gss(labels = percent,
                   breaks = seq(0, 1, 0.2),
                   limits = c(0, 1)) +
  scale_y_cont_gss(labels = c(paste0(enc2utf8("\u00A3"), "0-14,999"),
                              paste0(enc2utf8("\u00A3"), "15,000-19,999"),
                              paste0(enc2utf8("\u00A3"), "20,000-29,999"),
                              paste0(enc2utf8("\u00A3"), "30,000-39,999"),
                              paste0(enc2utf8("\u00A3"), "40,000-49,999"),
                              paste0(enc2utf8("\u00A3"), "50,000+"))) +
  labs(x = NULL,
       y = NULL,
       subtitle = "Estimated proportion of woodland visitors") +
  theme_gss(grid = "y",
            tick_marks = T)
```

### Supplementary analysis

Supplementary analyses were conducted to look at duration of woodland visit and socio-demographic profile, and main activity undertaken during woodland visit, both questions concerning the last fortnight. (This is in contrast to question M1_Q2_4 which asked respondents about woodland engagement in the last month).

-   Frequency of woodland visit was estimated by asking *How many times, if at all, did you make this type of visit to green and natural spaces in the last 14 days?* (No_of_Visits).

-   Activity was estimated by asking *Which of these activities, if any, did you do on this specific visit?* (M2A_Q8A).

### Supplementary Table 1: Tabulation of IMD decile and urban-rural classification {#sec-supplementary-table-1}

| IMD decile | Urban-rural classification | Sample size ($n$) |
|------------|----------------------------|-------------------|
| 1          | NA                         | 1,923             |
| 2          | NA                         | 1,968             |
| 3          | NA                         | 2,114             |
| 4          | NA                         | 2,110             |
| 5          | NA                         | 1,937             |
| 6          | NA                         | 1,995             |
| 7          | NA                         | 2,024             |
| 8          | NA                         | 2,019             |
| 9          | NA                         | 2,085             |
| 10         | NA                         | 1,907             |
| NA         | Rural                      | 1,420             |
| NA         | Urban                      | 3,364             |
| NA         | NA                         | 128               |

### Supplementary Table 2: Model performance comparison for full model including IMD decile or urban-rural classification {#sec-supplementary-table-2}

| Model                   | AIC      | Akaike weight ($w_i$) | Tjur's $R^2$ |
|-------------------------|----------|-----------------------|--------------|
| Model incl. IMD decile  | 20,206.8 | \<0.001               | 0.045        |
| Model incl. urban-rural | 4,060.0  | \>0.999               | 0.074        |

### Supplementary Table 3: Multi-model comparison

```{r}
#| echo: false
#| message: false
multi_model_summary <- readr::read_csv(here("data", "multi-model-comparison.csv"))

multi_model_summary %>% 
  filter(w > 0.01) %>% 
  mutate(aic = scales::comma(aic, accuracy = 1),
         w = round(w, 2),
         r2 = round(r2, 3)) %>% 
  knitr::kable(col.names = c("Formula", "AIC", "$w_i$", "Tjur's $R^2$"))
```

\*Showing models with $w_i > 0.01$ only.
