---
title: "An analysis of woodland engagement and socio-demographic profile in Wale's People and Nature Survey (2020/21)"
author: Daniel Braby
description: "Programme 3: Societal benefits of trees, woods and forests"
date: today
date-format: "D MMMM YYYY"
embed-resources: TRUE
format:
  html
editor: visual
editor_options: 
  chunk_output_type: console
---



```{r}
#| include: false

require(magrittr)

pans_y2 <- haven::read_sav("wales_21.sav") %>%
  janitor::clean_names()

pans_y2 <- pans_y2 %>%
  dplyr::select(m1_q2_4, m2a_q2, no_of_visits, age_band, qualification, income, home_imd_decile, ethnicity) %>%
  dplyr::rename(woodland_visit = m1_q2_4)


pans_y2 <- pans_y2 %>%
  dplyr::mutate(dplyr::across(-woodland_visit, ~ if (sjlabelled::is_labelled(.)) as.character(sjlabelled::as_label(.)) else .)) %>%
  dplyr::mutate(dplyr::across(-woodland_visit, as.factor)) 


m1_q2_4 <- pans_y2 %>% 
  dplyr::select(-m2a_q2,
                -no_of_visits) %>%
  dplyr::filter(!income %in% c("Don’t know", "Prefer not to say")) %>%
  dplyr::mutate(
    income = forcats::fct_relevel(
      income,
      "£0–14,999",
      "£15,000–19,999",
      "£20,000–29,999",
      "£30,000–39,999",
      "£40,000–49,999",
      "£50,000–59,999",
      "£60,000–79,999",
      "£80,000–99,999",
      "£100,000-£149,999",
      "£150,000 +"
    )
  )
  



## Breakdown of socio-demographic profile

create_summary_table <- function(data, var, target = "woodland_visit") {
  data %>%
    dplyr::group_by(.data[[var]], .data[[target]]) %>%
    dplyr::summarise(count = dplyr::n(), .groups = "drop") %>%
    dplyr::group_by(.data[[target]]) %>%
    dplyr::mutate(percent = count / sum(count)) %>%
    dplyr::mutate(label = paste0(count, " (", scales::percent(percent, accuracy = 1), ")")) %>%
    dplyr::select(category = .data[[var]], visited = .data[[target]], label) %>%
    tidyr::pivot_wider(names_from = visited,
                       values_from = label,
                       values_fill = "0 (0%)") %>%
    dplyr::rename(`Socio-demographic characteristic` = category,
                  `Did not visit wooland in last month` = `0`,
                  `Visited woodland in last month` = `1`)
    
}

```



# Methodology

## Study design and sample

Data were from year 2 of the [People and Nature Survey](https://beta.ukdataservice.ac.uk/datacatalogue/series/series?id=2000125) (PANS) for Wales (Natural Resources Wales, 2022). PANS is a nationally representative omnibus survey commissioned by Natural Resources Wales which gathers information on people's engagement, understanding and attitude towards the natural environment. Data were collected online (by Kantar). The survey builds on the Monitor of Engagement with the Natural Environment (MENE) survey which ran from 2009 to 2019.

## Data

Woodland engagement was estimated by asking *Which of the following type(s) of green and natural spaces have you visited during the last month?* (M1_Q2_4). Of the 5,954 respondents, 2,125 (36%) answered yes they had visited woodland or forest.

The following socio-demographic characteristics were controlled for in the analysis: age, education (level of highest qualification), employment status, ethnicity, Index of Multiple Deprivation (IMD) decile and income. Socio-demographic characteristics are summarised in Table 1.

#### Table 1: Breakdown of socio-demographic profile

```{r}
#| echo: FALSE
#| message: false
#| warning: false

create_summary_table(m1_q2_4, "age_band") %>%
  knitr::kable()

create_summary_table(m1_q2_4, "qualification") %>%
  knitr::kable()

create_summary_table(m1_q2_4, "income") %>%
  knitr::kable()

create_summary_table(m1_q2_4, "home_imd_decile") %>%
  knitr::kable()

create_summary_table(m1_q2_4, "ethnicity") %>%
  knitr::kable()



```


Analyses were conducted in R 4.2.1 (R Core Team, 2022). We used generalised linear modelling to investigate the association between woodland engagement (dependent/response variable) and socio-demographic profile (independent variables/predictors).


Significant socio-demographic characteristics were identified based on Analysis of Variance (ANOVA) using likelihood ratio chi-squared statistics (Fox & Weisberg, 2019). Estimated marginal means for each socio-demographic characteristic were estimated post-hoc using the `ggemmeans()` function in the `{ggeffects}` package (Lüdecke, 2018). Model fit statistics (AIC, $w_i$, Tjur's $R^2$) were calculated using the `{performance}` package (Lüdecke, 2021).

# Results

```{r}
#| include: false

y2_model <- glm(woodland_visit ~.,
                data = m1_q2_4,
                family = "binomial")

summary(y2_model)
```

Findings Summary

The logistic regression model examines the likelihood of a respondent visiting a woodland in the last month (Y = 1) based on socio-demographic and geographic predictors. Key findings are summarized below:

⸻

Significant Predictors

1. Age:
	•	Older adults (55–64, 65+): Strong negative association with woodland visits. Respondents in these age groups are significantly less likely to visit woodlands compared to the reference category (p < 0.001).
	•	Young adults (25–39): Showed a marginally positive trend (p = 0.082), suggesting higher odds of visiting compared to the baseline.

2. Qualifications:
	•	Respondents with no qualifications were significantly less likely to visit woodlands (\beta = -0.48, p = 0.003) compared to those with higher qualifications.

3. Income:
	•	Middle-income groups (e.g., £20,000–29,999, £30,000–39,999, £40,000–49,999, £50,000–59,999) were positively associated with woodland visits (p < 0.05).
	•	High-income groups (£80,000–99,999) also showed a strong positive association (\beta = 0.53, p = 0.006).
	•	Very high incomes (£150,000+) were not significantly associated.

4. Geographic Deprivation (IMD Decile):
	•	Living in less deprived areas (higher IMD deciles) positively correlated with woodland visits:
	•	IMD deciles 5–10 were all significantly associated with higher odds of visiting, with increasing strength as deprivation decreased (p < 0.01).

5. Ethnicity:
	•	Asian or Asian British (\beta = -2.10, p < 0.001) and Black or Black British (\beta = -1.38, p = 0.036) respondents were significantly less likely to visit woodlands.
	•	Mixed and White respondents showed no significant differences compared to the baseline.

⸻

Insignificant Predictors
	•	Income bands below £20,000 and above £100,000-£149,999 showed no significant impact.
	•	Some geographic IMD deciles (e.g., 2, 3, 4) and the qualifications category “Any other qualifications” were not statistically significant.

⸻

Model Fit
	•	Residual Deviance: 6015.6 (improved from null deviance of 6260.6, indicating the model explains some variance).
	•	AIC: 6073.6 (acceptable fit).
	•	Significant Predictors Count: Several predictors across demographics, income, area deprivation, and ethnicity were meaningful contributors.

⸻

Conclusion

The likelihood of visiting a woodland is influenced by a combination of socio-demographic factors (age, income, ethnicity, qualifications) and geographic characteristics (area deprivation). Middle-income groups, less deprived areas, and certain age bands (young adults) are more likely to visit, while older adults, less qualified individuals, and certain ethnic minorities are less likely to participate. These insights can inform targeted outreach or policy interventions to encourage more equitable woodland engagement.

#### Figure 1: Estimated marginal means for woodland engagement by age

```{r}
#| echo: false
ggeffects::ggemmeans(y2_model, "age_band", rg.limit = 272000) %>%
  dplyr::as_tibble() %>%
  dplyr::mutate(
    predicted = predicted * 100,  # Convert to percentages
    conf.low = conf.low * 100,    # Convert CI lower bound to percentages
    conf.high = conf.high * 100   # Convert CI upper bound to percentages
  ) %>%
  ggplot2::ggplot(ggplot2::aes(x = x, y = predicted)) +
  ggplot2::geom_point(size = 2) +
  ggplot2::geom_linerange(ggplot2::aes(
    ymin = conf.low,
    ymax = conf.high
  ), linewidth = 1) +
  ggplot2::scale_y_continuous(
    labels = scales::percent_format(scale = 1), # Format y-axis as percentage
    limits = c(0, 100)                         # Set y-axis limits to 0-100%
  ) +
  ggplot2::labs(
    x = NULL,
    y = "Percentage",
    subtitle = "Estimated proportion of woodland visitors"
  ) +
  afcharts::theme_af()
```

#### Figure 2: Estimated marginal means for woodland engagement by income

```{r}
#| echo: false
ggeffects::ggemmeans(y2_model, "income", rg.limit = 272000) %>%
  dplyr::as_tibble() %>%
  dplyr::mutate(
    predicted = predicted * 100,  # Convert to percentages
    conf.low = conf.low * 100,    # Convert CI lower bound to percentages
    conf.high = conf.high * 100   # Convert CI upper bound to percentages
  ) %>%
  ggplot2::ggplot(ggplot2::aes(y = x, x = predicted)) +  # Flip x and y
  ggplot2::geom_point(size = 2) +
  ggplot2::geom_linerange(ggplot2::aes(
    xmin = conf.low,  # Flip ymin to xmin
    xmax = conf.high  # Flip ymax to xmax
  ), linewidth = 1) +
  ggplot2::scale_x_continuous(
    labels = scales::percent_format(scale = 1), # Format x-axis as percentage
    limits = c(0, 100)                         # Set x-axis limits to 0-100%
  ) +
  ggplot2::labs(
    x = "Percentage",
    y = NULL,  # Remove y-axis label
    subtitle = "Estimated proportion of woodland visitors"
  ) +
  afcharts::theme_af()
```

#### Figure 3: Estimated marginal means for woodland engagement by home Index of Multiple Deprivation (IMD) decile

```{r}
#| echo: false
ggeffects::ggemmeans(y2_model, "home_imd_decile", rg.limit = 272000) %>%
  dplyr::as_tibble() %>%
  dplyr::mutate(
    predicted = predicted * 100,  # Convert to percentages
    conf.low = conf.low * 100,    # Convert CI lower bound to percentages
    conf.high = conf.high * 100   # Convert CI upper bound to percentages
  ) %>%
  ggplot2::ggplot(ggplot2::aes(y = x, x = predicted)) +  # Flip x and y
  ggplot2::geom_point(size = 2) +
  ggplot2::geom_linerange(ggplot2::aes(
    xmin = conf.low,  # Flip ymin to xmin
    xmax = conf.high  # Flip ymax to xmax
  ), linewidth = 1) +
  ggplot2::scale_x_continuous(
    labels = scales::percent_format(scale = 1), # Format x-axis as percentage
    limits = c(0, 100)                         # Set x-axis limits to 0-100%
  ) +
  ggplot2::labs(
    x = "Percentage",
    y = NULL,  # Remove y-axis label
    subtitle = "Estimated proportion of woodland visitors"
  ) +
  afcharts::theme_af()
```


#### Figure 4: Estimated marginal means for woodland engagement by ethnicity

```{r}
#| echo: false
ggeffects::ggemmeans(y2_model, "ethnicity", rg.limit = 272000) %>%
  dplyr::as_tibble() %>%
  dplyr::mutate(
    predicted = predicted * 100,  # Convert to percentages
    conf.low = conf.low * 100,    # Convert CI lower bound to percentages
    conf.high = conf.high * 100   # Convert CI upper bound to percentages
  ) %>%
  ggplot2::ggplot(ggplot2::aes(y = x, x = predicted)) +  # Flip x and y
  ggplot2::geom_point(size = 2) +
  ggplot2::geom_linerange(ggplot2::aes(
    xmin = conf.low,  # Flip ymin to xmin
    xmax = conf.high  # Flip ymax to xmax
  ), linewidth = 1) +
  ggplot2::scale_x_continuous(
    labels = scales::percent_format(scale = 1), # Format x-axis as percentage
    limits = c(0, 100)                         # Set x-axis limits to 0-100%
  ) +
  ggplot2::labs(
    x = "Percentage",
    y = NULL,  # Remove y-axis label
    subtitle = "Estimated proportion of woodland visitors"
  ) +
  afcharts::theme_af()
```
